Estimating duration in the presence of misclassification

### 2024-03-27 introduction_new.tex

\emph{Transmission} refers to the process by which an infectious individual generates new infections. Estimating transmission refers to inferring this process. Understanding components of transmission  (\it{e.g.} the number of contacts each individual makes per day) allow interventions to be targeted.
and to forecast future incidence.

\emph{Prevalence} is the proportion of the population with the disease at a given time .
Whether a given individual is \emph{prevalent}, {\it{i.e.}} has the disease, is often measurable through diagnostic testing.
Performing diagnostic tests in a sample of the population at a given point in time is known as a \emph{prevalence survey}.
Prevalence surveys can estimate the population prevalence but are typically very expensive to power appropriately.
Prevalence and incidence are strongly related, via the duration of an infection (see \cref{E-inc-prev}).

@@

Collecting data that {\it directly} inform inference of the quantities in this section is challenging or prohibitively expensive .
Hence, they are typically estimated from other data sources through appropriate inferential methodologies.{\color{red{refer to some papers examples:Phil O'neil, stats in medicine, 1990 on linking models to data; ; De Angelis, Presanis, 2018, Analysing Multiple Epidemic Data Sources. "Handbook of
Infectious Disease Data Analysis", Held, L., Hens, N., O'Neill, P.D. and Wallinga, J. (Eds.).
Chapman & Hall/CRC, 2019.}}}

@@

Data sources to inform outbreak surveillance are often observational in nature.
Therefore, they suffer from biases that can be hard to understand and quantify~\autocite{shadboltChallenges}.
These biases need to be accounted for when analysing the data.

@@

Incidence-type data originate from observations of the consequences of an infection, \eg the development of symptoms or admission to hospital.
This type of data typically suffers from reporting biases, particularly that the proportion of infections which are detected can vary over time or within different subgroups of the population~\autocites[chapter 9]{lashModern}{shadboltChallenges}.

An example of incidence-type data is the number of cases identified through community testing, reliant on individuals presenting for a test.
However, individuals' ability and willingness to be tested varies over time, which needs to be accounted for in order to perform unbiased inference.
Unfortunately, data to estimate these changes is normally unavailable.

Data on severe disease is typically less affected by these biases.
An extreme example is the development of AIDS following infection by HIV.
Here, the disease is so severe that it is extremely unlikely for an individual to develop the disease without seeking healthcare~\autocite{evansCompleteness}.

@@

Prevalence-type data originates from observations of the number or proportion of a population in a specific \emph{disease state} at a specific time.
For example, the number of individuals hospitalised with the disease or the proportion of the population with a marker of prior infection.
The issues associated with incidence-type data often also apply to prevalence-type data.

Prevalence surveys designed to draw random samples from the population are one form of prevalence-type data that can be unbiased.
Here, a representative proportion of the population (or a sub-population of interest) are tested for a marker of current infection.
For many diseases, prevalence is sufficiently low that this is prohibitively expensive to power.

However, \emph{seroprevalence} surveys are more feasible.
Seroprevalence is the proportion of individuals in the population which have antibodies to the disease, indicating that the individual was previously infected~\autocite{cdcSeroprevalence}.
Many more individuals have ever been infected than are currently infected, therefore, seroprevalence surveys can be powered much more cheaply~\autocite{wuSeroprevSimulation}.
However, seroprevalence is a delayed measure because there is a lag between an individual being infected and them developing antibodies.

\subsection{Cohort studies}

Other epidemiological and/or biological quantities are normally required to estimate transmission.
These depend on the model type, often including the duration of different disease states or the distribution of the time between infection events~\autocite{wallingaGI,dankwaStructural}.
Cohort studies (following a group of individuals over time), are a study design which can estimate many of these quantities.

Well-designed cohort studies can provide direct estimates of incidence and/or prevalence in the cohort~\autocite[chapter 7]{lashModern}.
However, recruiting a representative sample of the population is challenging, and the results of these studies are often not generalisable to the whole population.
Furthermore, these studies often require a large cohort to be appropriately powered.
Large cohort studies are expensive to run.

@@

The first component required to estimate incidence and transmission is a model of the infection process, a mathematical description of the process through which incidence arises.
A spectrum of models exists, categorised by the strength of assumptions they make about the transmission process~\autocite{beckerCOVIDmodels}.
Stronger assumptions allow more insight into the transmission process and lower the variance of the estimates but can introduce bias and be computationally challenging to fit.

The models with the weakest assumptions are \emph{phenomenological} models.
These make almost no assumptions about the transmission process, viewing incidence as a generic counting process (introduced in full in \cref{E-inc-prev:sec:infection-process}).
The most common model is that incidence follows a Poisson process, and the intensity of the process is the target of inference.
% The use of Poisson processes to model incidence of an infectious disease was very popular in the context of HIV/AIDS\todo{maybe a ref for Poisson in HIV/AIDS?}.

Next, are semi-mechanistic models.
These have components of the transmission process in them but do not fully represent the process.
A common type of semi-mechanistic model is those based on a \emph{renewal equation}~\autocite{bhattSemimechanistic}.
The renewal equation for an infectious disease~\autocite{fraserRenewal}, in the most general form, is:
\begin{align}
    Z(t) &= \int_{0}^{t} \beta(t,\tau) Z(t-\tau) d \tau
    \label{intro:eq:renewal}
\end{align}
where $Z(t)$ is the mean incidence at time $t$ and $\beta(t, \tau)$ is the mean rate at which an individual infected $\tau$ time units ago infects individuals at calendar time $t$.
Therefore, the renewal equation relates incidence in the past to incidence in the future.
This partially describes the transmission process, hence the name semi-mechanistic.
The effective reproduction number is $\Re = \int_0^\infty \beta(t, \tau) d\tau$.
% A scaled measure of $I(t)$ (\eg hospital admissions or community test results) allows estimating $\beta$.
% The most common implementations are the R packages EpiNow2~\autocite{EpiNow2} and EpiEstim~\autocite{EpiEstim}.
Extensions of \cref{intro:eq:renewal} that track the number of susceptible individuals become full mechanistic models, equivalent to the compartmental models introduced next~\autocite{champredonEquivalence}.
Renewal equations do not feature further in this thesis; for further details, including practical considerations when using them as the basis for inference, see \textcite{nashEstimating,gosticPractical}.

Finally, mechanistic models are the most complex models, representing the transmission process in varying levels of detail.
\emph{Compartmental} models, the most common type of mechanistic model, split the population into compartments, each representing a different state of the disease.
The \emph{dynamics} of the epidemic, how it changes over time, are represented by the transitions between compartments (for more details see \cref{E-SEIR:sec:mechanistic-models}).
Simple mechanistic models use a homogeneous population, with individuals differing only due to their disease state, and exponential distributions for the time in each disease state.
More complex models relax these assumptions, in particular to incorporate heterogeneity.
The age structure of the population is important for respiratory diseases because individuals meet individuals of their own age more frequently.
% Mathematically, the dynamics are described by ODEs (ordinary differential equations) or SDEs (stochastic differential equations).
% Unlike renewal equations, there are not widely-used implementations of compartmental models.
% A commonly used frameork is the SEIR (susceptible, exposed, infectious, recovered) model.
% The SEIR model, has four components (susceptible, exposed, infectious, recovered), with the population divided into these components; it is described in detail in \cref{E-SEIR:sec:mechanistic-models}.

\section{Disease models} \label{intro:sec:disease-model}

\emph{Disease models} link the transmission model to a latent quantity, $W_t$.
$W_t$ will then be recorded through some system (\eg a testing programme) described by the observation model (see next section).
For example, $W_t$ could be the incidence or prevalence of symptomatic individuals on day $t$.

At the core of many observation models is a \emph{convolution equation} or an approximation of it~\autocite[e.g.][]{brookmeyerBackcalculation,abbottEstimating,birrellRealtimea}.
This equation describes the expectation of $W_t$, conditional on the incidence prior to time $t$.
In discrete time, it is:
\begin{align}
    \E(W_t \mid Z_1, Z_2, \dots) &= \zeta \sum_{i=0}^\infty Z_{t-i} f(i)
    \label{intro:eq:disease-model}
\end{align}
where $Z_t$ is the incidence at time $t$; $\zeta$ is the proportion of infections that ever appear in $W_t$; and $f(i)$ is the probability that an infection at time $t-i$ is included in count $W_t$, conditional on it ever being in one of the counts.
$f$ is known as a \emph{delay function}.
$f$ is commonly the pdf of a distribution, or a transformation of it.

@@

\section{Observation models}

Observation models link $W_t$ to a possibly biased and/or noisy observation of it at time $t$.

Biases can arise in a variety of ways, \eg healthcare-seeking behaviour or tests which are less than 100\% accurate.
Bias can be modelled through additional parameters or data quantifying these issues~\autocite[e.g.][]{sherrattExploring,birrellRealtimea,nicholsonImproving,swallow2022challenges}.

Noise could come from a variety of sources, \eg sampling error, clustering, or unexplained overdispersion.
Noise is typically modelled by a parametric distribution \eg the negative binomial distribution allowing for overdispersion compared to a Poisson process assumption~\autocite[e.g.][]{birrellRealtimea,frassoBayesian}.

@@

Applying \cref{intro:eq:disease-model} requires the delay function, $f$.
Typically, this is either the time until an event or the duration of a disease stage.
Time series data on $W_t$ cannot normally identify $f$, instead external information needs incorporating~\autocite{swallow2022challenges}.

Two approaches are commonly used to estimate the duration of a disease stage: either modelling an underlying biomarker or modelling the duration directly~\autocite[e.g.][]{sweetingEstimating}.
Modelling a biomarker uses more information per observation but requires stronger assumptions.

The biomarker approach models the trajectory of a biomarker over time in a cohort of individuals.
An individual is in the disease state while the biomarker is above or below some threshold.
This approach often requires parametric assumption(s) such as a functional form of the biomarker's trajectory.
In addition, if individual-level variation is included, which is generally important, a distributional form of random effects is required.
This requires both sufficient observations of the biomarker per individual and an assumption on the form of the distribution.
More details are in \cref{E-ATACCC}.

Modelling the duration directly considers only the binary result at each test (positive or negative), for instance in a \emph{survival analysis} framework.
This approach generally requires fewer parameters reducing the computational burden.
It is also easier to specify non-parametric forms of the distribution, which is important for SARS-CoV-2 as the tail behaviour may be important.
% A more detailed review of the literature is in the relevant chapters.
% Survival analysis is the area of statistics concerned with estimating the distribution of the times between two events.
% Other related approaches, such as a multi-state model (reviewed by \textcite{jacksonMSM}), could also be considered; multi-state models can be viewed as a generalisation of survival analysis.
More details are in \cref{E-perf-test}.

Both cases require longitudinal data, suggesting that cohort studies are the most appropriate study design.

@@

The surveillance systems used in response to SARS-CoV-2 are historically unique.
Modern technology allowed large amounts of data to be collected, distributed, and analysed.
While the HIV/AIDS pandemic inspired the development of a large amount of statistical methodology, it was far slower moving.
Much analysis of SARS-CoV-2 data was done at great haste, requiring trade-offs to be made between timeliness and rigour of results presented.

A deep understanding of each data source's potential was not possible in real-time.
In this thesis, I will investigate the potential of the prevalence surveys in greater detail.
The disease model will be a prevalence model (defined further in \cref{E-inc-prev:sec:prevalence-process}) with the delay function being the duration of PCR positivity.

@@

The main data sources informing the response to the SARS-Cov-2 epidemic in England were based on: community testing (known as pillar 2 testing), the number of severe events, seroprevalence surveys, and prevalence surveys.
Each of these data sources have strengths and weaknesses, see \textcite{royalSocietyRnumber} and references therein for further details on each of these data sources.

@@

For large parts of the pandemic, community testing was widely available in the UK.
However, it did not test a representative sample of the population.
The number of tests conducted varied over time and space, for a variety of known and unknown reasons.
These include: the policy on who is eligible for testing, the availability of tests, and the willingness of individuals to be tested.
Therefore, fluctuations in the number of positive tests may not reflect changes in the true incidence of the disease.
The large number of tests conducted allowed detailed analyses to be performed, including stratification by small geographic areas or age groups, although this does not solve the issue of the biases present.

Severe events counts are generally a more reliable measure of the burden of the disease.
The criteria for being a severe event are normally well-defined and generally well-recorded in the UK context.
Therefore, these data are subject to fewer biases than community testing.
However, SARS-CoV-2 rarely leads to severe events in younger individuals, meaning severe event counts are insensitive to changes in transmission only affecting younger individuals.

@@

Seroprevalence surveys were conducted throughout the epidemic, the largest of which was tested donated blood~\autocite{amirthalingamSeroprevalence}.
In addition to the normal problem of the delay between infection and antibody response, blood donors are a convenience sample whose representativeness of the wider population is unclear.

The extraordinary impact of SARS-CoV-2 led the UK Government to fund two large prevalence surveys in representative samples of the general population.
These were designed to avoid the biases of community testing and observe the whole population.
Therefore, some modelling is required to borrow information across strata or time to provide stable estimates, as well as to correct for non-response bias (see \cref{E-intro:sec:cis}).
Despite this, they are considered the gold-standard metric.

The use of prevalence surveys for estimating incidence and transmission is the focus of this thesis.
These surveys use RT-PCR (reverse transcription polymerase chain reaction, described in detail in \cref{E-biology-data:sec:PCR}) testing to detect the presence of the virus in a swab from an individual.
I focus on the CIS (Coronavirus (COVID-19) Infection Survey), run by the ONS (Office for National Statistics)~\autocite{CIS,cisMethodsONS}.
The CIS was a longitudinal study, testing a large number of individuals weekly or monthly for SARS-CoV-2; it is described in detail in \cref{E-intro:sec:cis}.

A second prevalence survey ran in the UK, known as REACT (REal-time Assessment of Community Transmission)~\autocite{rileyResurgence,rileyREACT}.
REACT was a series of cross-sectional surveys, run across 19 rounds between May 2020 and March 2022.
Each round consisted of at least \numprint{100000} randomly sampled individuals from the UK contributing swabs for testing and questionnaire data~\autocite{elliottTwin}.

A major advantage of the CIS over REACT is that the CIS had continual data collection, while REACT has no test results between rounds.
The lack of data between rounds meant that using REACT data to infer transmission in these periods requires strong model assumptions.
A particular instance of this is that a peak in late December 2020 occurred between rounds of the REACT survey, complicating analysis of the survey data at that time~\autocite{rileyREACTround8}.

\subsubsection{Cohort studies}

Many cohort studies, to estimate a wide variety of quantities, ran throughout the pandemic.
In this thesis, I make use of two cohort studies with differing designs to estimate the duration of PCR positivity.
These designs have complementary strengths, and I combine information from the studies.
Both of the studies recruited individuals from the general population, aiming to observe a cohort of infections that are representative of the infections in the UK community.

% In addition to population-level surveillance, cohort studies help understand other aspects of the pathogen~\autocite{royalSocietyRnumber}.
% For example, contact tracing data can inform how likely it is for transmission to occur in different settings and/or between different groups of people.
The first study I used, ATACCC (Assessment of Transmission and Contagiousness of COVID-19 in Contacts)~\autocite{singanayagamCommunity,hakkiOnset}, performed daily testing of a small number of individuals with short follow-up.
The density of this sampling allows for reliable estimates of the viral load dynamics and other quantities of interest, most importantly the duration of detectability, but the results are limited because the small sample size and short follow-up prohibit estimation of the tails of these distributions.
I describe ATACCC in detail in \cref{E-biology-data:sec:ataccc} and use it to estimate duration in \cref{E-ATACCC}.

@@

Members of SPI-M-O (the Scientific Pandemic Influenza Group on Modelling, Operational sub-group) produced the most prominent modelling of the UK epidemic.
SPI-M-O was a sub-committee of SAGE (the Scientific Advisory Group for Emergencies), generating evidence based on infectious disease modelling for policy development~\autocite{medleySPIM,govSPIMO}.
This used a wide range of  modelling approaches depending on the data available, which fluctuated throughout the epidemic, and the question being addressed.
Some were fit to data while others were purely used for simulation based on plausible parameter values.

Each week, members of SPI-M-O would produce a set of estimates of the current state of the epidemic (\eg the value of $\Re$), known as \emph{nowcasting}.
The committee peer-reviewed these estimates to provide quality assurance, removing any estimates unduly affected by issues (\eg data quality issues).
They then formed a consensus value using a statistical combination of the estimates~\autocite{parkCombining}.
The consensus value was published on the UK Government's website and informed its response to the epidemic~\autocite{govRnumber}.

The nowcasting estimates were based on mechanistic and semi-mechanistic models with a wide variety of data sources.
The mechanistic models commonly incorporated multiple data sources, known as \emph{data synthesis}, a technique reviewed by \textcite{birrellEvidence}.
The models used were reviewed by \textcite{royalSocietyRnumber} (see appendix 2 of that review for details of the specific implementations).

@@

In this thesis, I consider how prevalence surveys could be used for epidemic surveillance, in particular estimates of incidence and transmission.
These surveys have well-understood designs which means that estimates derived from them are more robust.
A crucial quantity in their use will be the distribution of the duration of RT-PCR positivity.

In the following subsection, I introduce the most relevant related work, from both prior to the pandemic and during its course.

@@

\emph{Backcalculation} is a modelling framework developed in the context of HIV/AIDS~\autocite{brookmeyerMethod} but applied very broadly.
It specifies a particular combination of infection, disease, and observation model which relate times of HIV infection with times of AIDS diagnosis via the incubation period.
Further details are in the recent review \textcite{sunModeling}.
The theory behind the framework underpins much of the work in this thesis.

The original formulations~\autocite{brookmeyerMethod,rosenbergBackcalculation} used a Poisson process as the infection model; a disease model where all infections are eventually observed ($\zeta = 1$) and the delay function, $f$, is the incubation period distribution's pdf; and an observation model where all diagnoses are observed, and hence has no influence on the data.

The disease and observation models must be adapted to apply backcalculation to SARS-CoV-2 prevalence surveys (see \cref{E-inc-prev}).
The disease model must consider that this is now prevalence-type, rather than incidence-type data.
This difference induces a dependency structure between the observations on each day.
The observation model must now include the with error arising from using a sample of the population rather than (assumed) full recording of AIDS diagnoses.

@@

An early meta-analysis estimated that individuals test positive for SARS-CoV-2 for a mean of 14.6 days (95\% CI: 9.3--20.0 days)~\autocite{cevikShedding}.
The longest reported duration was observed at 83 days, although beyond day 9 no live virus was detected, indicating that the individuals testing positive for longer than this were not infectious.
However, most studies included in the meta-analysis involved hospitalised patients or had unclear inclusion criteria, and hence may not be representative of the general population.
These summary estimates are insufficient for backcalculation, which requires a distribution of the duration of positivity.

\Textcite{hellewellPCRSensitivity} estimated the probability of an individual testing positive each day after infection.
This measure is highly related to the duration of positivity.
However, the study's cohort consisted only of healthcare workers, \ie working-age adults.

\Textcite{binnySensitivity} applied a slight variant of \textcite{hellewellPCRSensitivity}'s methodology to data from New Zealand.
The situation in New Zealand, who were successfully pursuing a zero-COVID strategy at this time, means that the individuals in the study were unusual.
For example, a third were travellers detected during their quarantine period on arrival to New Zealand but would have had negative RT-PCR tests before their departure.

>>>>

\emph{Transmission} refers to the process by which an infectious individual generates new infections. Estimating transmission refers to inferring this process. Understanding components of transmission  {\it{e.g.}} the number of contacts each individual makes per day) allow interventions to be targeted.
and to forecast future incidence.

\emph{Prevalence} is the proportion of the population with the disease at a given time .
Whether a given individual is \emph{prevalent}, {\it{i.e.}} has the disease, is often measurable through diagnostic testing.
Prevalence surveys, designed  to perform diagnostic testing of a sample of the population,  can be used to estimate the population prevalence,  but are typically very expensive to power appropriately.
Prevalence and incidence are strongly related, via the duration of an infection (see \cref{E-inc-prev}).

@@

Collecting data that {\it directly} inform inference of the quantities in this section is challenging or prohibitively expensive .
Hence, they are typically estimated from other data sources through appropriate inferential methodologies.
{\color{red}{refer to some papers examples:Phil O'neil, stats in medicine, 1990 on linking models to data; De Angelis, Presanis, Analysing Multiple Epidemic Data Sources. "Handbook of
Infectious Disease Data Analysis", Held, L., Hens, N., O'Neill, P.D. and Wallinga, J. (Eds.).
Chapman & Hall/CRC, 2019.}}

@@

Data sources to inform outbreak surveillance are often observational in nature, {\it i.e.} they are not the output of designed studies.
Therefore, they suffer from biases that can be hard to understand and quantify~\autocite{shadboltChallenges}.
These biases need to be accounted for when analysing the data.

@@

Incidence-type data originates from observations of the consequences of an infection, \eg the development of symptoms or admission to hospital.
This type of data typically suffers from reporting biases, as the proportion of infections which are detected can vary over time or within different subgroups of the population~\autocites[chapter 9]{lashModern}{shadboltChallenges}.

An example of incidence-type data is the number of cases identified through community testing, which relies on individuals presenting for a test.
Individuals' ability and willingness to be tested varies over time, which needs to be accounted for in order to perform unbiased inference.

Data on severe disease is typically less affected by biases.
An extreme example is the development of AIDS following infection by HIV.
Here, the disease is so severe that it is extremely unlikely for an individual to develop the disease without seeking healthcare~\autocite{evansCompleteness}.

@@

Prevalence-type data derives from observations of the number or proportion of a population in a specific \emph{disease state} at a specific time.
For example, the number of individuals in hospitalwith the disease or the proportion of the population with a marker of prior infection.
The issues associated with incidence-type data often also apply to prevalence-type data.

Prevalence surveys are one source of prevalence-type data that can be unbiased.
For instance, a representative proportion of the population (or a sub-population of interest) might be tested for a marker of current infection.
For many diseases, prevalence is sufficiently low that this is prohibitively expensive to power.

On the other hand, \emph{seroprevalence} surveys are more feasible.
Seroprevalence is the proportion of individuals in the population which have antibodies against a particular infection, indicating that the individual was previously infected~\autocite{cdcSeroprevalence}.
Many more individuals have ever been infected than are currently infected, therefore, seroprevalence surveys can be powered much more cheaply~\autocite{wuSeroprevSimulation}.
However, seroprevalence is a delayed measure because there is a lag between an individual being infected and them developing antibodies.

\subsection{Other studies}

Other epidemiological and/or biological quantities are normally required to estimate transmission.
These include the duration of different disease states or the distribution of the time between infection events~\autocite{wallingaGI,dankwaStructural}.
Specific studies, {\it e.g.} longitudinal studies that follow  individuals over time, allow  estimation of estimate many of these quantities.

Well-designed cohort studies can provide direct estimates of incidence and/or prevalence in the cohort~\autocite[chapter 7]{lashModern}.
However, recruiting a representative sample of the population is challenging, and the results of these studies are often not generalisable to the whole population.
Furthermore, these studies often require a large cohort to be appropriately powered, which are expensive to run.

@@

The first component required to estimate incidence and transmission is a model of the infection process, a mathematical description of the process through which infections are generated.
A spectrum of models exists, categorised by the strength of assumptions they make about the transmission process~\autocite{beckerCOVIDmodels}.

The models with the weakest assumptions are the so-called \emph{phenomenological} models (add a reference!! Still Becker?).
These make almost no assumptions about the transmission process, viewing incidence as resulting from a point process (introduced in full in \cref{E-inc-prev:sec:infection-process}).
A common model is the Poisson process, and the intensity of the process is the target of inference.
% The use of Poisson processes to model incidence of an infectious disease was very popular in the context of HIV/AIDS\todo{maybe a ref for Poisson in HIV/AIDS?}.

Next, are semi-mechanistic models.
These include components of the transmission process  but do not explicitly describe the process.
A common type of semi-mechanistic models is that based on a \emph{renewal equation}~\autocite{bhattSemimechanistic} of the most form:
\begin{align}
    Z(t) &= \int_{0}^{t} \beta(t,\tau) Z(t-\tau) d \tau
    \label{intro:eq:renewal}
\end{align}
where $Z(t)$ is the expected incidence at time $t$ and $\beta(t, \tau)$ is the expected rate at which an individual infected $\tau$ time units ago infects individuals at calendar time $t$.
Therefore, the renewal equation relates incidence in the past to incidence in the future.
This partially describes the transmission process, hence the name semi-mechanistic.
The effective reproduction number is $\Re = \int_0^\infty \beta(t, \tau) d\tau$.
% A scaled measure of $I(t)$ (\eg hospital admissions or community test results) allows estimating $\beta$.
% The most common implementations are the R packages EpiNow2~\autocite{EpiNow2} and EpiEstim~\autocite{EpiEstim}.
Renewal equations do not feature further in this thesis; for further details, including practical considerations when using them as the basis for inference, see \textcite{nashEstimating,gosticPractical}.

Finally, mechanistic models are the most complex models, representing the transmission process in varying levels of detail.
\emph{Compartmental} models, the most common type of mechanistic model, split the population into compartments, each representing a different state of the disease.
The \emph{dynamics} of the epidemic are represented by the transitions between compartments (for more details see \cref{E-SEIR:sec:mechanistic-models}).
Simple mechanistic models use a homogeneous population, with individuals differing only due to their disease state, and exponential distributions for the time in each disease state.
More complex models relax these assumptions to incorporate heterogeneity, for example, to account for the age structure and the interaction between age groups.
% Mathematically, the dynamics are described by ODEs (ordinary differential equations) or SDEs (stochastic differential equations).
% Unlike renewal equations, there are not widely-used implementations of compartmental models.
% A commonly used frameork is the SEIR (susceptible, exposed, infectious, recovered) model.
% The SEIR model, has four components (susceptible, exposed, infectious, recovered), with the population divided into these components; it is described in detail in \cref{E-SEIR:sec:mechanistic-models}.

\section{Observational models} \label{intro:sec:disease-model}
 The infection process is hardly ever directly observed. However, as seen in previous sections,  indirect observations are available on the infection's sequelae, which can be linked to the latent infection process through appropriate observational models.
At the core of many observation models is a \emph{convolution equation} or an approximation of it~\autocite[e.g.][]{brookmeyerBackcalculation,abbottEstimating,birrellRealtimea}.
This equation describes the expectation of an infectious sequelae process ({\it e.g.} doctor's consultation, death) $W_t$, conditional on the incidence prior to time $t$.
In discrete time, it is:
\begin{align}
    \E(W_t \mid Z_1, Z_2, \dots) &= \zeta \sum_{i=0}^\infty Z_{t-i} f(i)
    \label{intro:eq:disease-model}
\end{align}
where $Z_t$ is the incidence at time $t$; $\zeta$ is the proportion of infections that result in $W_t$; and, conditional on being observed, $f(i)$ describes the (discretised) probability distribution of the time from infection to observation.
$f$ is an example of a \emph{delay function}, and  
$f$ is commonly the pdf of a distribution, or a transformation of it.

@@

@@

Applying \cref{intro:eq:disease-model} requires knowledge of the delay function, $f$.
Typically, this is either the time to an event or the duration of a disease stage.
Time series data on $W_t$ cannot identify $f$, instead external information needs incorporating~\autocite{swallow2022challenges}.  This generally comes from longitudinal studies where the occurrence of the infection's sequelea is monitored in a cohort of individuals ({\it e.g.} Collaborative Group on AIDS Incubation and HIV Survival including the CASCADE EC
Concerted Action (2000). Time from HIV- 1
seroconversion to AIDS and death in 13,000 individuals before the widespread use of
highly active anti-retroviral therapy: effect of age at seroconversion and exposure
category. Lancet 355, 1131-1137.)

Two approaches are commonly used to estimate the duration of a disease stage: either modelling an underlying biomarker or modelling the duration directly, for instance, in a \emph{survival analysis} framework.~\autocite[e.g.][]{sweetingEstimating}.

The biomarker approach models the trajectory of a biomarker over time.
An individual has an observable/detectable infection while the biomarker is above or below some threshold.
This approach often requires parametric assumption(s) such as a functional form of the biomarker's trajectory.
In addition, if individual-level variation is included, which is generally important,  sufficient observations of the biomarker per individual and an assumption on the form of the distribution are required
More details are in \cref{E-ATACCC}.

Modelling the duration directly considers only the binary result at each test (positive or negative).
This approach generally requires fewer parameters reducing the computational burden.
It is also easier to specify non-parametric forms of the distribution, which is useful in situation where the tail behaviour may be important.
% A more detailed review of the literature is in the relevant chapters.
% Survival analysis is the area of statistics concerned with estimating the distribution of the times between two events.
% Other related approaches, such as a multi-state model (reviewed by \textcite{jacksonMSM}), could also be considered; multi-state models can be viewed as a generalisation of survival analysis.
More details are in \cref{E-perf-test}.

@@

The response to the SARS-CoV-2 pandemic in the UK has been unprecedented, both in terms of the scientific input to policy and the surveillance data made available. Members of the Scientific Pandemic Influenza Group on Modelling Operational sub-group (SPI-M-O ), a sub-committee of the Scientific Advisory Group for Emergencies (SAGE), generated evidence throughout the pandemic to inform policy development~\autocite{medleySPIM,govSPIMO}.
Each week, SPI-M-O would produce a set of estimates of the current state of the epidemic (\eg the value of $\Re$), known as \emph{nowcasting}.
The committee peer-reviewed these estimates to provide quality assurance and formed a consensus value using a statistical combination of the estimates~\autocite{parkCombining}.
The consensus value was published on the Government's website and informed its response to the pandemic~\autocite{govRnumber}.

The nowcasting estimates were based on mechanistic and semi-mechanistic models with a wide variety of data sources.
The mechanistic models, in  particular, commonly incorporated multiple data sources \textcite{birrellEvidence}.
The models used were reviewed by \textcite{royalSocietyRnumber} (see appendix 2 of that review for details of the specific implementations).

@@

The surveillance systems used in response to SARS-CoV-2 are historically unique. 
The main data sources informing the response to the SARS-Cov-2 epidemic in England were based on: community testing; severe events; seroprevalence and prevalence surveys.
Each of these data sources have strengths and weaknesses, see \textcite{royalSocietyRnumber} and references therein for further details on each of these data sources.

@@

For large parts of the pandemic, community testing was widely available in the UK.
However,
the number of tests carried out varied over time and space, for a variety of reasons, including: policy changes on who was eligible for testing;availability of tests; and willingness of individuals to be tested.
Therefore, fluctuations in the number of positive tests did not necessarily reflect changes in disease incidence.
The large number of tests conducted allowed detailed analyses to be performed, including stratification by small geographic areas or age groups, although the lack of interpretability remained.

Severe events counts were generally a more reliable measure of the burden of the disease.
The criteria for being a severe event are normally well-defined and generally well-recorded.
Therefore, these data were subject to fewer biases than the cases detected through community testing.
However, SARS-CoV-2 rarely led to severe events in younger individuals, so severe event counts were insensitive to changes in transmission only affecting younger individuals.

@@

However, some modelling is required to borrow information across strata or time to provide stable estimates, as well as to correct for non-response bias (see \cref{E-intro:sec:cis}).
Despite this, they are considered the gold-standard metric.

Seroprevalence surveys were conducted throughout the epidemic, the largest of which was testing donated blood~\autocite{amirthalingamSeroprevalence}.
In addition to the problem of the delay between infection and antibody response, blood donors are a convenience sample whose representativeness of the wider population is unclear.
The extraordinary impact of SARS-CoV-2 led the UK Government to fund two large prevalence surveys in representative samples of the general population.
These were designed to avoid the biases of community testing and attempt to observe the general population.
These surveys use RT-PCR (reverse transcription polymerase chain reaction, described in detail in \cref{E-biology-data:sec:PCR}) testing to detect the presence of the virus in a swab from an individual.
The first survey was the  Coronavirus (COVID-19) Infection Survey (CIS), run by the Office for National Statistics (ONS)~\autocite{CIS,cisMethodsONS}.
The CIS was a longitudinal household study, testing a large number of individuals weekly or monthly for SARS-CoV-2 infection (see \cref{E-intro:sec:cis} for a detailed description).

A second prevalence survey, known as the REal-time Assessment of Community Transmission (REACT)~\autocite{rileyResurgence,rileyREACT},
consisted in a series of cross-sectional surveys, run across 19 rounds between May 2020 and March 2022.
Each round included at least \numprint{100000} randomly sampled individuals from the UK contributing swabs for testing and questionnaire data~\autocite{elliottTwin}.

A major advantage of the CIS over REACT is that the CIS had continual data collection, while REACT has no test results between rounds.
This meant that using REACT data to infer transmission in these periods required strong model assumptions.
A particular instance of this is that a peak in transmission in late December 2020 occurred between rounds of the REACT survey, challenging analysis of the REACT data at that time~\autocite{rileyREACTround8}.

\subsubsection{Other studies}

Many ad hoc studies to estimate a wide variety of epidemiological and biological quantities of the SARS-CoV2 virus were carried out throughout the pandemic.
% In addition to population-level surveillance, cohort studies help understand other aspects of the pathogen~\autocite{royalSocietyRnumber}.
% For example, contact tracing data can inform how likely it is for transmission to occur in different settings and/or between different groups of people.
One of the study, which  I will use in what follows, is the Assessment of Transmission and Contagiousness of COVID-19 in Contacts (ATACCC) study. ~\autocite{singanayagamCommunity,hakkiOnset}. InATACCC  daily testing of a small number of individuals with short follow-up was performed.
The density of this sampling allows for reliable estimates of the viral load dynamics and other quantities of interest, most importantly the duration of detectability. However, the results from ATACCC are of limited value because of the small sample size and, particularly, of the short follow-up, which  prohibit estimation of the tails of these distributions.
I describe ATACCC in detail in \cref{E-biology-data:sec:ataccc} and use it to estimate duration in \cref{E-ATACCC}.

@@

@@

During the SARS-CoV-2 pandemic, analysis of the unprecedently available data was carried out in haste, as results were needed at pace. Therefore, a deep understanding of the potential of each data source was not possible.
In this thesis I investigate the potential of the CIS as a sinlge source of information for incidence and transmission estimation.

In the following subsection, I introduce the most relevant related work, carried out  both prior to the pandemic and during its course.

@@

\emph{Backcalculation} is a modelling framework developed in the context of HIV/AIDS~\autocite{brookmeyerMethod} but applied very broadly.
It specifies a particular combination of infection and observation model which relate times of HIV infection with times of AIDS diagnosis via the incubation period.
Further details are in the recent review \textcite{sunModeling}.
The theory behind the framework underpins much of the work in this thesis.

The original formulations~\autocite{brookmeyerMethod,rosenbergBackcalculation} uses a Poisson process as the infection model and a delay function, $f$, the distribution of the time between infection and symptomatic disease (incubation period); and an observation model where all diagnoses are observed.

In (see \cref{E-inc-prev}) I will adapt this backcalculation framework to data from the CIS, which is of a prevalence-type rather than incidence-type.
This difference induces a dependency structure between the observations on each day.
The observation model must now include the with error arising from using a sample of the population rather than (assumed) full recording of AIDS diagnoses.

@@

In the application of the backcalculation to CIS data, the delay function $f$ is represented by the duration of positivity. Many attempts have been made to estimate its distribution. An initial meta-analysis estimated that individuals test positive for SARS-CoV-2 for a mean of 14.6 days (95\% CI: 9.3--20.0 days)~\autocite{cevikShedding}, with 
the longest reported duration at 83 days, also reporting lack of virus detection beyond day 9, indicating that the individuals testing positive for longer than this were not infectious.
However, most studies included in the meta-analysis involved hospitalised patients or had unclear inclusion criteria, and hence may not be representative of the general population.
In addition, summary estimates are insufficient for backcalculation, which requires a distribution of the duration of positivity.

\Textcite{hellewellPCRSensitivity} estimated the probability of an individual testing positive each day after infection.
This measure is highly related to the duration of positivity.
However, the study's cohort consisted only of healthcare workers, {it i.e.} working-age adults.

\Textcite{binnySensitivity} applied a slight variant of the methodology  in \textcite{hellewellPCRSensitivity} to data from New Zealand.
The situation in New Zealand, who were successfully pursuing a zero-COVID strategy in the earlier part of the pandemic, means that the individuals in the study were unusual.
For example, a third were travellers detected during their quarantine period on arrival to New Zealand but would have had negative RT-PCR tests before their departure.
